Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-02-15 src/article.tex
Typo in figure 1 legend
Revise manuscript

\author{
  Michael D Barton$^{1}$%
  \email{Michael D Barton - mail@michaelbarton.me.uk}%
\and
  Hazel A Barton\correspondingauthor$^1$%
  \email{Hazel A Barton\correspondingauthor - bartonh@nku.edu}%
      }

@@

  \paragraph*{Background:} Assembly of short read sequencing data can result in
  a fragmented series of contigs. Therefore a common step in a genome project
  is to join neighbouring sequence regions together and fill gaps using insert
  data generated through PCR. This scaffolding step, however, is non-trivial
  and requires manually editing large blocks of nucleotide sequence. Joining
  sequence contigs together also hides the source of each region in the final
  genome sequence. Taken together, these considerations may make reproducing or
  editing an existing genome build difficult.

  \paragraph*{Methods:} The software outlined ``Scaffolder'' is implemented in
  the Ruby programming language and can be installed via the RubyGems software
  management system. Genome scaffolds are defined using YAML - a markup
  language, which is both human and machine-readable. Command line binaries and
  extensive documentation are provided.

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent contigs using a relatively simple to write and edit
  scaffold file syntax. The Scaffolder syntax further allows inserts to be
  added to the scaffold for the purpose of filling unknown sequence regions.
  Defining the genome construction in a file makes the scaffolding process
  reproducible and comparatively easier to edit than raw fasta.

@@

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each read represents the machine-interpreted
nucleotide sequence from a short region in the genome. At the time of writing
high throughput sequencing is limited to producing reads less than $>$1000
nucleotides. Therefore to produce to produce a sequence longer than this
length, e.g. a complete genome, these numerous individual reads must be pieced
together. The process of joining reads into longer sequences is the `assembly'
stage in a genome project \cite{assembly-reviews}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence compose of
these individual reads. An analogy to describe this is a jigsaw puzzle. Each
nucleotide read represents a single piece and the final genome sequence is the
completed puzzle. Repetitive nucleotide `repeat' regions within a genome or
biased and incomplete sequencing data may however prevent the genome being
assembled into a continuous sequence. This is analogous to missing jigsaw
pieces or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has lead
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{assembly-reviews}. Genome assembly software is often unable to
completely these sequence reads together due to problems outlined above.
Instead assembly algorithms will generate large contiguous regions of sequence
(`contigs') composed of many individual reads. These represent optimum assembly
of reads into the longest sequences for the assembly algorithm. \pb

The process of ameliorating a draft sequence to a finished continuous sequence
can be expensive in time and laboratory effort. Therefore in many cases a set
of contigs may often describe enough of the genomic content for many research
questions \cite{branscomb2002}. A continuous high-quality `finished' genome
sequence does however provide further depth of information
\cite{parkhill2002,fraser2002}. The benefits of a complete genome sequence
include complete resolution of repeat regions and the exact distances between
genomic elements. This process of joining contigs from assembly together to form
a continuous sequence may be called `scaffolding' or `finishing'. Here on we
use the term scaffolding. \pb

@@

Scaffolding is the process of joining a series disconnected contigs into
complete continuous sequence. Due to genomic complexity, missing data and
diminishing returns scaffolding may not ultimately produce a final completed
sequence. The process of scaffolding may still however joining a subset of
contigs or additional regions of sequence to fill gaps.

@@

The sequencing process produces reads from either strand of the DNA helix. The
resulting contigs constructed from these reads therefore point in either of two
DNA strand directions. Contig orientation involves reverse complementing
sequences so that all contigs point in the same direction and represent the
same DNA strand. In the case of microbes this orientation is often the 5`-3`
strand leading away from the origin of replication.

@@

Contig ordering determines the placement of contigs to best represent the
actual nucleotides as appearing in the genome sequence. This may also determine
if any contigs represent extra-genomic DNA, such as plasmids, and therefore
should be treated separately from genomic contigs. The contig order is commonly
referenced to the origin of replication where the contig containing the
replication start point is first. All subsequent contigs then follow in the
direction of DNA replication.

@@

Given the order and orientation of contigs determining the distances between
contigs may be useful for estimating genome size. The size of each gaps between
contigs represents the length of an unknown region in the genome. Filling in
these regions in with unknown nucleotide characters (`N') allows a draft
continuous representation of the genome to produced. This draft scaffold
sequence represents both the contigs and unknown regions of sequence between
them.

\subsubsection*{Gap Closing and Filling} %{{{3

During the scaffolding process closing and filling sequence gaps improves the
accuracy of the genome scaffold. This may require returning to the laboratory
to perform additional sequencing or using computational methods. Once all
contigs have been joined and gaps in a scaffold closed the genome may be
considered as finished. In the case of organisms or particular interest
additional sequencing may be performed to further `proof' any low coverage
regions or detect erroneous nucleotides resulting from insertions, deletions or
substitutions.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or both in conjunction.
Examples of computation methods is the use of pair reads data, where sequencing
reads are a known distance apart in the genome, or from comparing contigs to
a reference genome. Laboratory methods may use PCR to amplify the DNA in gap
regions, then use traditional Sanger sequencing to determine this sequence.
Computation methods are more preferable as they less costly in laboratory time
and materials compared to manual gap resolution \cite{nagarajan2010}. \pb

The need for computational scaffolding methods has resulted in a variety of
algorithmic approaches. Synteny methods compare the assembled contigs to a more
complete reference genome sequence. Areas of corresponding sequence between the
the two sets are indicative of the possible contig organisation
\cite{richter2007,zhao2008}. \pb

Repeat regions may be responsible for many of the gaps in building a genome
sequence. The existence of tandemly repeated nucleotide regions produces many
reads of similar sequence. The assembly of these regions may result in the
repeats being collapsed into a single instance or ignored by conservative
assembly algorithms. These regions may be reassembled using algorithms
specifically for repeat resolution \cite{mulyukov2002,koren2010}. A related
approach uses uniquely matching reads try find overlaps and bridge across the
gaps between contigs \cite{tsai2010}. \pb

As mentioned above, the genome may be prepared prior to sequencing so that
reads are produced in pairs at a known distance apart. This paired-read data
provides an extra level of information on how contigs may be scaffolded
together. Heuristic graph algorithms may be used to search for optimal
configuration of contigs in the scaffold to take into account this information
\cite{dayarian2010}. Additional data using a reference genome may also be
combined when estimating the best contig configuration \cite{pop2004}. \pb

@@

Taken together these methods provide a wide variety of possible approaches for
scaffolding a genome. Scaffolding however may still however require manual
editing of the scaffold if computational methods do not resolve the final
sequence.  This process involves moving and joining larges blocks of nucleotide
text together. This however may be error-prone and manual editing also
precludes reproducibility of the genome scaffold. \pb

The software outlined here ``Scaffolder'' is aimed at creating a syntax and
software framework for editing a genome scaffold. Scaffolder uses a specific
data syntax to define how contigs are joined, additional sequences inserted,
and specify unknown regions. This syntax allows a scaffold to be edited in much
easier way and makes genome scaffolds reproducible. \pb

@@

Scaffolder is written in the Ruby programming language, version 1.8.7
\cite{ruby-lang}. The scaffolder package is split into two libraries. The first
called ``scaffolder'' contains the core scaffolder application programming
interface (API). The second library ``scaffolder-tools'' contains the
scaffolder command line interface (CLI). \pb

Unit tests were implemented to develop and maintain individual elements of the
source code.  Integration tests were written to test the scaffolder software
interface as a whole. Unit tests were written using the Shoulda and RSpec
\cite{rspec} libraries.  Integration tests were written using the Cucumber
library \cite{rspec}. \pb

The Scaffolder source code is documented using the Yard library \cite{yard}.
Unix manual pages for the command line interface were generated using the Ronn
library \cite{ronn}. The manipulation of biological sequences in Scaffolder is
performed using the BioRuby library \cite{goto2010}. A full list of the
software dependencies in scaffolder can be found in the Gemfile in the root of
each source code directory.\pb

@@

The choice of nucleotide sequences making up the scaffold is specified using
the YAML syntax \cite{yaml}. YAML is a language format using whitespace and
indentation to produce a machine readable structure. As YAML is a standardised
data format, third-party developers have the option to generate a genome
scaffold using any programming language for which a YAML parser and generator
exists. The YAML website lists parsers for languages including C/C++, Ruby,
Python, Java, Perl, C\#/.NET, PHP, and JavaScript. In addition to being widely
supported YAML formatted scaffold files can be validated for correct syntax
using third party tools such as Kwalify \cite{kwalify}. \pb

As described in the Background section, the initial sequencing data assembly
may result in an incomplete genome build. The use of computational software and
generation of additional sequence using PCR translates into genome scaffolding
as an on-going process. Therefore the scaffold file should be simple to edit
update with new sequence manually in addition to being computationally
tractable. This requirement was therefore second reason YAML was selected,
where YAML is human-readable and should be relatively easy to update in a
standard text editor. \pb

The scaffold file takes the form of a list of entries. Each entry corresponds
to a region of sequence, most likely a contig, in the genome scaffold. The
scaffolder software processes this list of entries in order, converting each to
the corresponding nucleotide sequence. These nucleotide sequences are then
concatenated together to produce the draft genome sequence. Each entry in the
scaffold file may have attributes that define whether a sub-sequence or the
reverse complement of the sequence should be used. The types of attributes
available and an example scaffold file are outlined in the Results and
Discussion section below.  \pb

@@

The Scaffolder software outlined here is designed to allows manual or
computational joining of nucleotide sequences together into a single contiguous
super sequence. Plain-text files written using the YAML syntax files are used
to specify the nucleotide sequences in the scaffold and how they should be
joined together. This file is described as the ``scaffold file''. \pb

The scaffold file is read by the scaffolder software to determine how the
scaffold sequence should be generated. In addition to specifying which
nucleotides should be used to generate this draft sequence, the scaffold file
allows the sequences to trimmed down to smaller sub-sequences and reverse
complemented if necessary. \pb

The process of genome finishing may involve producing additional nucleotide
sequence to fill gaps in the scaffold. The scaffold file format allows these
additional insert sequences to be added to scaffold to bridge the gaps inside
contigs. These inserts can also be treated in a similar way to larger contig
sequences: trimmed and/or reverse complemented to match the corresponding gap
region. \pb

In the process of build a genome scaffold distances between contigs may be
estimated from paired read insert size or from mapping the contigs to reference
genome. These unknown gap regions are useful to join non-contiguous sequences
together by the estimated size of the region whilst still highlighting areas in
the scaffold to be resolved. The scaffold file also allows for the
specification of these unknown regions. \pb

The scaffold file specifies how the contigs nucleotide sequences are joined
together, however the sequences are not present in the scaffold file. These
nucleotide sequences are instead maintained in a fasta file. A fasta file of
contigs is a common output for assemblers and as such should be readily
available without further processing. Nucleotide sequence are specified in the
scaffold using the first space-delimited word from the fasta header.
Maintaining the nucleotide sequences in a separate file preserves the original
unedited sequence and furthermore separates raw sequence from the markup
describing how it should be used in the scaffold. \pb

\subsection*{Defining a scaffold in a text file} %{{{2

The scaffold file is written using the YAML syntax. YAML is a data format
designed to be both machine readable and easy to edit manually. An example
scaffold file is shown in Figure 1. This file illustrates the text attributes
used to describe a genome scaffold. The basic layout of the scaffold file is as
a list - a series of entries where each entry insert to a region of sequence in
the scaffold super sequence. \pb

@@

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. Line 2 begins
with a dash character `-', dashes are used to denote entries in a YAML list.
This first line therefore indicates indicate the first entry in the scaffold.
Line 3 is indented by two spaces where whitespace is used to group similar
attributes together. Lines indented by whitespace are linked to the first
non-indented preceding line. \pb

The ``sequence'' tag on line 3 indicates this scaffold entry corresponds to an
entry in the fasta file. The following line, indented by two spaces, indicates
the name of the fasta sequence using the ``source'' keyword. As described above
the first space-delimited word of the fasta header should be used to identify
which sequence is used. Therefore lines 2-4 describe the first entry in the
scaffold as a simple sequence using the fasta sequence named `sequence'. \pb

@@

Line 5 has no whitespace indentation and begins with a dash therefore
describing the next entry in the scaffold. This entry is identified by the
``unresolved'' tag, and specifies a region of unknown nucleotides but known
length. The following line specifies the size of this unknown region. Lines 5-6
therefore insert a region of 20 `N' characters in the scaffold. \pb

@@

The last entry adds the fasta entry `sequence2' to the scaffold. This entry
demonstrates how sequence may be edited when added to the scaffold. The `start'
and `stop' tags trim the sequence to these coordinates inclusively. The
``reverse'' tag reverse complements this region. \pb

This region further includes the ``inserts'' tag to update the inserted
sequence with additional regions of sequence. These inserts are also added as
a YAML list with each insert listed with a dash on one line followed by the
remainder of the insert errata. \pb

Lines 16-19 illustrate an insert using similar attributes to that of a sequence
entry. The reverse, start and stop tags trim the inserted sequence and the
`source' tag identifies the corresponding fasta sequence. The ``open'' and
``close'' tags on lines 20-21 determine where the insert is added to the
enclosing sequence.  These coordinates inclusively replace the sequence region
within these coordinates with the specified insert fasta sequence. The open and
close positions need not be the same size as the replaced region. \pb

The next insert on lines 23-24 is specified by only the open position tag. This
illustrates that only the one of the `open' or `close' tags is required.  When
only one of these tag is used the corresponding opposing coordinate position is
calculated from the length in the insert fasta sequence. This allows inserts to
bridge into and partially fill gap regions. \pb

@@

The sequence entry in the scaffold file allows multiple sequence-editing
operations to be applied. If applied in different combinations, these
operations could result in different output sequence. An example would be
reverse complementing the sequence then trimming the start of the sequence by
five nucleotides. Performing these actions in the reverse order would produce
a difference output sequence. This section therefore outlines the scaffolder
protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts these are reverse sorted by the insert close coordinate. The insert
with the largest close position is first in the list and the insert with the
smallest close position is last. This preserves the original indented
coordinates for each insert during the process of adding inserts to the
sequence. \pb

\paragraph{Update sequence with insert:} Each insert sequence is processed
according to any specified attributes. The insert is then added to the
enclosing sequence in the region specified by the open and close coordinates.
\pb

If the insert added to the sequence is a different size compared to the region
it replaces then the sequence start and stop co-ordinates are updated.  For
example if a 7bp insert replaced a 5bp region in the sequence the stop
coordinate of the enclosing sequence would be incremented by the 2bp
difference. As inserts are added by reverse close position order the remaining
inserts are unaffected by this change in the sequence size. \pb

\paragraph{Trimming sequence:} If the sequence has start or stop sequences
specified these are used to trim the sequence to the specified length
inclusively. \pb

\paragraph{Reverse complement sequence:} If the reverse tag is specified the
sequence is reverse complemented. This is final step in processing the sequence
and results in the sequence being returned for inclusion in the scaffold. \pb

\subsection*{Scaffolder usage} %{{{2

Scaffolder provides a standardised set of Ruby classes and methods (API)
allowing the scaffold to be manipulated programmatically. This allows
scaffolder to be integrated into existing genomics workflows or used with Ruby
build tools such as Rake. \pb

In addition Scaffolder provides a command line interface (CLI) through the
scaffolder binary. This allows the scaffold sequence to be produced and
validated without requiring any Ruby experience. The scaffolder CLI behaves as
a standard Unix tool - returning appropriate exit statuses and providing
appropriate manual pages. The use of both these scaffolder interfaces is
outlined in the following sections. \pb

@@

If both these commands return without error this indicates the necessary
system requirements to install and run scaffolder. The scaffolder API and CLI
can then be installed through RubyGems with the following command. \pb

@@

To ensure that the scaffolder CLI is accessible, the RubyGems `bin' directory
should be in the command line path. Assuming that RubyGems are installed in
~/.gem this can be done by adding the following to the ~/.bashrc file. The
corresponding login shell should be updated for other shells such as Zsh, etc..
\pb

@@

The Scaffolder CLI provides functions to validate the scaffold file and build
the draft scaffold sequence. Assuming scaffolder has been installed and the
binary is available in the command line path the following command should run
without error and list the available scaffolder functions. \pb

@@

Help on a specific command can be found by typing `help' followed by the name
of the command. This returns the corresponding Unix manual page. The following
example will return the manual page for the sequence command. \pb

@@

The ``sequence'' command builds the fasta sequence from the scaffold and fasta
file. Running this command requires the locations of both the scaffold file and
the fasta file to be passed as argument. This will then return the complete
scaffold sequence in fasta format. \pb

@@

Scaffolder provides a function to test the scaffold for overlapping insert
coordinates. Overlapping inserts will result in unexpected behaviour as the
coordinates for the first insert may change the insert location for subsequent
inserts. The ``validate'' command will check the scaffold for any such inserts
and return them as a YAML list. \pb

@@

\subsubsection*{Application programming interface} %{{{3

Scaffolder provides an API allowing Ruby programs to manipulate the scaffold.
This API is detailed in full on the scaffolder website. The following code
snippet outlines how the API may be used to create a script to replicate the
function of the `scaffolder sequence' command described above in the CLI
section. \pb

@@

    require 'rubygems'
    require 'scaffolder'

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable object and the next line illustrates the
trivial code required to generate the scaffold sequence. Further details of the
classes and method related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class. \pb

@@

MDB developed and maintains the scaffolder tool. MDB and HAB wrote the
manuscript.

>>>>

\author{
  Michael D Barton$^{1}$\correspondingauthor%
  \email{Michael D Barton\correspondingauthor  - mail@michaelbarton.me.uk}%
\and
  Hazel A Barton$^1$%
  \email{Hazel A Barton - bartonh@nku.edu}%
      }

@@

  \paragraph*{Background:} Assembly of short read sequencing data can result
  in a fragmented series of contigs. Therefore a common step in a genome
  project is to join neighbouring sequence regions together and fill gaps
  using additional sequence. This scaffolding step, however, is non-trivial
  and requires manually editing large blocks of nucleotide sequence. Joining
  sequence contigs together also hides the source of each region in the final
  genome sequence. Taken together, these considerations may make reproducing
  or editing an existing genome build difficult.

  \paragraph*{Methods:} The software outlined here ``Scaffolder'' is
  implemented in the Ruby programming language and can be installed via the
  RubyGems software management system. Genome scaffolds are defined using
  YAML - a markup language, which is both human and machine-readable. Command
  line binaries and extensive documentation are available.

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent contigs using a relatively simple to write and edit
  scaffold file syntax. This syntax further allows  unknown regions to be
  defined, and inserts to be added to the scaffold to fill contig gaps.
  Defining the genome construction in a file makes the scaffolding process
  reproducible and comparatively easier to edit and compare than raw fasta.

@@

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each read represents the machine-interpreted
nucleotide sequence from a short region in the genome. At the time of writing,
high throughput sequencing is limited to producing reads less than $>$1000
nucleotides long. Therefore a sequence longer than this length, e.g.
a complete genome, these numerous individual reads must be pieced together.
The process of joining reads into longer sequences is the `assembly' stage in
a genome project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence compose of
these individual reads. An analogy for this process is a jigsaw puzzle. Each
nucleotide read represents a single piece and the final genome sequence is the
completed puzzle. Repetitive nucleotide `repeat' regions within a genome or
biased and incomplete sequencing data may however prevent the genome being
assembled into a continuous sequence. This is analogous to missing jigsaw
pieces or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has lead
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}. Current genome assembly software is however
often unable to produce a complete sequence from these reads due to these
problems. In many cases assembly algorithms will generate several large
contiguous regions of sequence (`contigs') composed of many individual reads.
These contigs represent the assembly of reads into the longest possible
sequences. These contigs represent a fragmented picture of the genome and
require additional work to piece together. These contigs represent
a fragmented picture of the genome and require additional work to piece
together. \pb

The process of ameliorating a draft sequence to a finished continuous sequence
can be expensive in time and laboratory effort. Therefore in many cases a set
of contigs may often describe enough of the genomic content for many research
questions \cite{branscomb2002}. A continuous high-quality `finished' genome
sequence does however provide further depth of information including complete
resolution of repeat regions and the exact distances between genomic elements
\cite{parkhill2002,fraser2002}. This process of joining contigs together to
form a continuous sequence may be called the `scaffolding' or `finishing'
stage. \pb

@@

Scaffolding is the process of joining a series disconnected contigs into
complete continuous sequence. Due to genomic complexity and missing data,
scaffolding may not ultimately produce a final completed sequence. The process
of scaffolding may still however succeed in joining a subset of contigs
together or using additional regions of sequence to fill gaps. An overview of
the scaffolding process is outlined in the following section.

@@

The sequencing process produces reads from either strand of the DNA helix.
Therefore the resulting contigs constructed from these reads may point in
either direction of the DNA strand. Contig orientation reverse complements
sequences where necessary so that all contigs point in the same direction and
represent the same DNA strand. In the case of microbes this orientation may be
the 5'-3' strand leading away from the origin of replication.

@@

Contig ordering determines the placement of contigs to best represent their
order in the true genome sequence. Correctly placing each contig may also
highlight any extra-genomic DNA, such as plasmids, and which should be treated
separately from genomic sequence. The contig order is commonly referenced
against the origin of replication where the contig containing the replication
start point is first. All subsequent contigs are then ordered in the direction
of DNA replication.

@@

Given the order and orientation of contigs, determining the correct distances
between contigs leads to an estimate of the genome size. The size of inter
contig gaps represents the length of an unknown region in the genome. Filling
in these regions in with unknown nucleotide characters (`N') allows a draft
continuous representation of the genome. This draft scaffold sequence is
useful for representing both the sequenced and unknown regions in the genome.

\subsubsection*{Gap Closing} %{{{3

During the scaffolding process closing and filling sequence gaps improves the
accuracy of the genome scaffold. This may require returning to the laboratory
to perform additional sequencing or using computational methods of estimating
the gap sequence. These additional sequences can be used to bridge and close
the gaps between contigs. Once all contigs have been joined and gaps in
a scaffold closed the genome may be considered finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both.
An example of a computational method uses pair read data, where the sequencing
reads are a known distance apart in the genome, when paired reads are in
separate contigs this can to estimate the order and distance between them.
Laboratory methods may use PCR to amplify the unknown DNA in gap regions,
followed by traditional Sanger sequencing. Computation methods are more
preferable as they less costly in laboratory time and materials compared to
manual gap resolution \cite{nagarajan2010}. The need for computational
scaffolding methods has resulted in a variety of algorithmic approaches. \pb

Synteny methods use sequence similarity to compare the assembled contigs to
a more complete reference genome sequence. Areas of corresponding sequence
between the the two sets are indicative of the possible contig organisation in
the true genome \cite{richter2007,zhao2008}. \pb

Repeat regions may be responsible for many of the gaps in building a genome
sequence. The existence of tandemly repeated nucleotide regions produces many
reads of similar sequence. The assembly of these regions may result in the
repeats being collapsed into a single instance or ignored by conservative
assembly algorithms. These regions may be reassembled using algorithms
specifically for repeat resolution \cite{mulyukov2002,koren2010}. A related
approach uses uniquely matching reads to try and bridge across contig gaps
between contigs \cite{tsai2010}. \pb

As mentioned above, the genome may be prepared prior to sequencing so that
reads are produced in pairs at a known distance. This paired-read data
provides an extra level of information on how contigs may be scaffolded
together. Heuristic graph algorithms may take advantage of this data to search
for optimal configuration of contigs in the scaffold to take this into account
\cite{dayarian2010}. Additional data using a reference genome may also be
combined when estimating the best contig configuration \cite{pop2004}. \pb

@@

Taken together these methods provide a wide variety of possible approaches for
scaffolding a genome. The scaffold may however still require manual editing
and addition of sequence if computational methods are not completely
successful. This process involves moving and joining larges blocks of
nucleotide text together. This process may be expected to be error-prone and
precludes any reproducibility of the scaffold. \pb

The software outlined here ``Scaffolder'' is aimed at creating a syntax and
software framework for producing and editing a genome scaffold. Scaffolder
uses a specific data syntax to define how contigs are joined, additional
sequences inserted, and for the specification of unknown regions. This syntax
allows a scaffold to be edited in a simpler way and allows the scaffolding
process to be reproducible and deterministic. \pb

@@

Scaffolder is written in the Ruby programming language, version 1.8.7
\cite{ruby-lang}. The scaffolder package is split into two libraries. The first
called ``scaffolder'' contains the core Scaffolder application programming
interface (API). The second library ``scaffolder-tools'' contains the
Scaffolder command line interface (CLI). \pb

Unit tests were implemented to develop and maintain the individual elements of
the source code.  Integration tests were written to test the Scaffolder
software interface as a whole. Unit tests were written using the Shoulda and
RSpec \cite{rspec} libraries.  Integration tests were written using the
Cucumber library \cite{rspec}. \pb

The Scaffolder source code is documented using the Yard library \cite{yard}.
Unix manual pages for the command line interface were generated using the Ronn
library \cite{ronn}. The manipulation of biological sequences in Scaffolder is
performed using the BioRuby library \cite{goto2010}. A full list of the
software dependencies in Scaffolder can be found in the Gemfile in the root of
each source code directory. \pb

@@

The choice of nucleotide sequences comprising the scaffold is specified using
the YAML syntax \cite{yaml}. YAML is a language format using whitespace and
indentation to produce a machine readable structure. As YAML is a standardised
data format, third-party developers have the option to generate a genome
scaffold using any programming language for which a YAML parser and generator
exists. The YAML website lists current parsers for languages including C/C++,
Ruby, Python, Java, Perl, C\#/.NET, PHP, and JavaScript. In addition to being
widely supported YAML formatted scaffold files can be validated for correct
syntax using third party tools such as Kwalify \cite{kwalify}. \pb

As described in the Background section, the initial sequencing data assembly
may result in an incomplete genome build. The use of computational software and
generation of additional sequence using PCR translates into genome scaffolding
as an on-going process. Therefore the scaffold file should be simple to edit
update manually in addition to being computationally tractable. This
requirement was therefore the second reason YAML was selected, where YAML is
human-readable and should be relatively easy to manipulate in a standard text
editor. \pb

The scaffold file takes the form of a list of entries. Each entry corresponds
to a region of sequence, most likely a contig, in the genome scaffold. The
Scaffolder software processes this list of entries in order, converting each to
the corresponding nucleotide sequence. These nucleotide sequences are then
concatenated together to produce the draft super sequence. Each entry in the
scaffold file may have attributes that define whether a sub-sequence or the
reverse complement of the sequence should be used. The types of attributes
available and an example scaffold file are outlined in the Results and
Discussion section below.  \pb

@@

The Scaffolder software outlined here facilitates or computational joining of
nucleotide sequences together into a single contiguous super sequence.
Plain-text scaffold files written in the YAML syntax are used to specify how
nucleotide sequences should be joined together. This scaffold file is read by
the software to determine how the draft scaffold sequence should be generated.
In addition to specifying which contigs should be used, the specification of
the scaffold file allows the contigs to trimmed down to smaller sub-sequences
and reverse complemented if necessary. \pb

The process of genome finishing may involve producing additional
oligonucleotide sequence to fill areas of unknown sequence in the scaffold.
The scaffold file format allows these additional insert sequences to be added
and bridge gaps. These inserts can also be treated in a similar way to larger
contig sequences: trimmed and/or reverse complemented to match the
corresponding gap region. \pb

In the process of building a genome scaffold distances between contigs may be
estimated from paired read data or from mapping the contigs to a reference
genome. These inter contig gap regions are useful to join regions of sequence
together by an estimated size. The scaffold file also allows for the
specification of these unknown regions to highlight areas in the genome
sequence still to be resolved. \pb

The scaffold file is used to specify how sequence regions are joined together,
however the sequences themselves are not stored in the scaffold file. The
sequences are instead maintained in a fasta file. A fasta file of contigs and
scaffolded contigs is a common output for assemblers and as such should be
readily available without further processing required. Nucleotide sequence are
reverenced in the scaffold using the first space-delimited word from each fasta
header. Maintaining the nucleotide sequences in a separate file preserves the
original sequence and furthermore decouples the raw sequence from the markup
describing how it should be used to generate the genome sequence. \pb

\subsection*{Defining a Scaffold in a Text File} %{{{2

The scaffold file is written using the YAML syntax. YAML is a data format
designed to be both machine readable and easy to edit manually. An example
scaffold file is shown in Figure 1. This file illustrates the text attributes
used to describe a scaffold and the corresponding genome build. The basic
layout of the scaffold file is a list - a series of entries where each entry
corresponds to a region of sequence in the scaffold super sequence. \pb

@@

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. The first
entry in the scaffold begins on Line 2. The first line of this entry,
highlighted in green, begins with a dash character `-', dashes are used to
denote entries in a YAML list. This first line therefore indicates indicate
the first entry in the scaffold. The first line of each entry in the scaffold
begins with a dash line. The second line of this entry indented by two spaces
where whitespace is used to group similar attributes together. Lines indented
by whitespace are linked to the first non-indented preceding line in the file.
\pb

The ``sequence'' tag indicates this entry corresponds to a sequence in the
fasta file. The following line indicates the name of the fasta sequence using
the ``source'' tag. As described above the first space-delimited word of the
fasta header is used to identify which sequence. Therefore these three lines
describe the first entry in the scaffold as a simple sequence using the fasta
sequence named `sequence1'. \pb

@@

The second entry in the scaffold, highlighted in orange, is identified by the
``unresolved'' tag, and specifies a region of unknown nucleotides but known
length. The following line in the entry specifies the size of this unknown
region. This entry therefore inserts a region of 20 `N' characters in the
scaffold. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the fasta entry
`sequence2' to the scaffold. This entry demonstrates how the nucleotide
sequence may be manipulated prior to addition to the scaffold. The `start' and
`stop' tags trim the sequence to these coordinates inclusively. The
``reverse'' tag reverse complements the sequence. \pb

This region further includes the ``inserts'' tag to update this sequence with
regions of sequence. These inserts are also added as a YAML list with each
insert beginning with a dash line. The first insert, shown in purple,
illustrates using similar attributes to that of a sequence entry. The reverse,
start and stop tags, are used to trim and reverse complement the insert.
Similarly the `source' tag identifies the corresponding fasta sequence. The
``open'' and ``close'' tags determine where the insert is added in the
enclosing sequence. These coordinates inclusively replace the sequence region
within these coordinates with the specified insert fasta sequence. \pb

The next insert, shown in brown, specifies only the open position tag. This
illustrates that only one of either `open' or `close' tags is required. When
only one of these tag is used the corresponding opposing coordinate position
is calculated from the length in the insert fasta sequence. This allows
inserts to bridge into, and partially fill gap regions. \pb

@@

The sequence entry in the scaffold file allows multiple sequence-editing
operations to be applied. If applied in different combinations, these
operations could result in different output sequence. An example would be
reverse complementing the sequence then trimming the start of the sequence by
five nucleotides. Performing these actions in the reverse order would produce
a difference output sequence. This section therefore outlines the Scaffolder
protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts, these inserts are reverse sorted by the insert close coordinate. The
insert with the largest close position is first in the list and the insert
with the smallest close position is last. This preserves the original intended
coordinates for each insert during the process of adding each insert to the
sequence. \pb

\paragraph{Update sequence with inserts:} Each insert sequence is processed
according to any specified attributes. The insert is then added to the
enclosing sequence in the region specified by the open and close coordinates.
\pb

If the insert added to the sequence is a different size to that of the
replaced region then the sequence start and stop co-ordinates are updated. For
example if a 7bp insert replaced a 5bp region in the sequence the stop
coordinate of the enclosing sequence would be incremented by the 2bp
difference. As inserts are added by reverse close position order the remaining
inserts are unaffected by changes in the encoding sequence size. \pb

\paragraph{Trimming sequence:} If the sequence has start or stop tags
specified these are used as coordinates to trim the sequence to the specified
length inclusively. \pb

\paragraph{Reverse complement sequence:} If the reverse tag is specified the
sequence is reverse complemented. This is final step in processing the
sequence and the sequence is then returned for inclusion in the scaffold. \pb

\subsection*{Scaffolder Usage} %{{{2

Scaffolder provides a standardised set of Ruby classes and methods (API) for
manipulation of the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. \pb

In addition Scaffolder provides a command line interface (CLI). This allows
the scaffold sequence to be generated and validated without any Ruby
experience. The Scaffolder CLI behaves as a standard Unix tool - returning
appropriate exit codes and providing manual pages. The use of both these
Scaffolder interfaces is outlined in the following sections. \pb

@@

If both these commands return without error this indicates the necessary
components are available to install and run Scaffolder. The Scaffolder API and
CLI can then be installed using the following command. \pb

@@

To ensure that the Scaffolder CLI is accessible, the RubyGems `bin' directory
should be in the command line path. Assuming that RubyGems are installed in
\~/.gem this can be done by adding the following to the \~/.bashrc file. The
corresponding login script should be updated for other shells such as Zsh,
etc.. \pb

@@

The Scaffolder CLI provides functions to validate the scaffold file and build
the draft scaffold sequence. Assuming Scaffolder has been installed and the
binary is available in the command line path the following command should run
without error. This will list the available Scaffolder functions. \pb

@@

Help on a specific command can be found by typing ``help'' followed by the
name of the command. This prints the corresponding Unix manual page to the
terminal. The following example will return the manual page for the sequence
command. \pb

@@

The ``sequence'' command builds the fasta sequence from the scaffold and fasta
file. Running this command requires the locations of both the scaffold file
and the fasta file to be passed as arguments. This will then return the
complete scaffold sequence in fasta format. \pb

@@

Scaffolder provides a function to test the scaffold for overlapping insert
coordinates. Overlapping inserts will result in unexpected behaviour when
generating the scaffold sequence as the coordinates for the first insert may
change the insert location for subsequent inserts. The ``validate'' command
will check the scaffold for any such inserts and return them as a YAML list.
\pb

@@

\subsubsection*{Application Programming Interface} %{{{3

Scaffolder provides an API allowing Ruby programs to manipulate the scaffold.
This API is detailed in full on the Scaffolder website. The following code
snippet outlines how the API may be used to create a script to replicate the
function of the `scaffolder sequence' command described in the CLI section
above. \pb

@@

    require `rubygems'
    require `scaffolder'

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable array and the next line illustrates the
trivial code required to generate the scaffold sequence. Further details of
the classes and method related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class on the Scaffolder website. \pb

@@

MDB developed and maintains the Scaffolder tool. MDB and HAB wrote the
manuscript.
